# Results `r Sys.Date()`

```{r set, include=FALSE, message = FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = F)
```

## Line-transect surveys 

```{r load_data}
{year <- 2022
  month <- "01"
  month_abb <- month.abb[as.numeric(month)]
  survey_title <- paste(cemore::first_up(month_abb), year) 
  folder = paste0("../survey_data/raw_data/",year, "-", month)
  surveyid = paste0("cemore_", year, tolower(month_abb))
  iterations <- data.frame(cbind(
    year = c(rep(2020,3),rep(2021, 11),rep(2022,1)),
    month_abb =c("Sep", "Oct", "Nov", "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Oct", "Nov", "Dec","Jan"),
    iteration = c(1,2,3,6,7,8,9,10,11,12,13,14,15,16,18)))
  iteration <- iterations$iteration[which(iterations$year == year & iterations$month_abb == month_abb)]
}

# Load previously saved data
all_effort <- load_effort(year, as.numeric(month), single_survey = F, vessel = "MB")
all_effort_lines <- readRDS(paste0("C:/users/keppele/documents/github/cemore/tech_report/output/all_effort_lines_to ", survey_title, ".rds"))
all_ap_sf <- readRDS(paste0("C:/users/keppele/documents/github/cemore/tech_report/output/all_effort_sgt_to ", survey_title, ".rds")) %>% 
mutate(Count =case_when(
  BestNumber == 1 ~ "1",
  BestNumber %in% c(2:5) ~ "2:5",
  BestNumber >5 ~ ">5"
) %>% factor(levels = c("1", "2:5", ">5")))
# filter for species of interest
# all_ap_sf %<>% filter(Species %like% "Humpback")
# total number of survey days and total count cetacean sightings; manually check any NA values


```

Total distance surveyed = `r round(sum(st_length(all_effort_lines))/1000) %>% as.numeric()` km

### Map seasonal tracklines to date - Jan-Mar, Apr-Jun-, Jul-Sep, Oct-Dec 

```{r plot_seasonal_tracklines, echo=F, include = F}

  
g <- plot_survey(Save = F,
                 border = T,
  single_survey <- F,
  incidentals <-  F,
  hydrophone <- F,
  sightings_only <- F)
```
```{r plot_tracklines}
g
```

### Summarize surveys
```{r summary_tables}
x <- all_effort_lines %>% 
  mutate(length = as.numeric(st_length(geometry))) %>% 
  as.data.frame() %>% filter(!is.na(length)) %>% 
  mutate(Month = month_abb, Year = year,Season=season) %>% dplyr::select(-c(month,year,season)) 
  # transmute(Year = year, Season = season, length = as.numeric(length)) %>%
 
x %>%  dplyr::group_by(Year, Month, Season) %>%
  dplyr::summarise("Total survey distance (km)" = round(sum(length/1000))) %>%
  csasdown::csas_table()

x %>% dplyr::group_by(Season) %>%
  dplyr::summarise("Total survey distance (km)" = round(sum(length/1000)),
                   "# Surveys" = n_distinct(SurveyID)) %>% 
  csasdown::csas_table()
```

### Survey Sightings
```{r survey_sightings_sum}
all_ap_sf %>%  dplyr::group_by(Species) %>%
  dplyr::summarise( N = n(), "# individuals" = sum(BestNumber),
                    "Mean grp size" = round(mean(BestNumber), 1),
                    sd = round(sd(BestNumber), 1),
                    "Std.err" = round(sd/sqrt(N),1),
                    "Min grp size" = round(min(BestNumber),1),
                    "Max grp size" = round(max(BestNumber),1)
                    ) %>%
                    as.data.frame() %>%
  dplyr::rename("# sightings" = N) %>% 
  dplyr::select(-c(sd,geometry)) %>% 
  csasdown::csas_table()
```

### Covariates

Survey distace each covariate was experienced

```{r covariate_plots}
all_effort_lines_sum <- all_effort_lines %>% mutate(length_km = as.numeric(st_length(geometry))/1000) %>% 
  as.data.frame() %>% mutate(Bf = factor(Bf, levels = c(0,1,2,3,4,5,6,7)))
suppressMessages(ggplot()) +
  geom_col(data = all_effort_lines_sum, aes(Bf, length_km)) #+
suppressMessages(ggplot()) +
  geom_col(data = all_effort_lines_sum, aes(Port.Vis, length_km))
suppressMessages(ggplot()) +
  geom_col(data = all_effort_lines_sum, aes(Swell, length_km))
suppressMessages(ggplot()) +
  geom_col(data = all_effort_lines_sum, aes(Glare, length_km))

```

Number of sightings at covariate levels

`r table(all_ap_sf$beauf) %>% data.frame() %>% csasdown::csas_table(col_names = c("Beaufort","Count"))` 
`r table(all_ap_sf$SightedBy) %>% data.frame() %>% csasdown::csas_table(col_names = c("Observer","Count"))` 
`r table(all_ap_sf$swell) %>% data.frame() %>% csasdown::csas_table(col_names = c("Swell","Count"))` 
`r table(all_ap_sf$port_visib) %>% data.frame() %>% csasdown::csas_table(col_names = c("Visibility","Count"))` 
`r table(all_ap_sf$glare) %>% data.frame() %>% csasdown::csas_table(col_names = c("Glare","Count"))` 

```{r BestNumber}
x <- all_ap_sf %>% dplyr::mutate(
  Count = factor(
    dplyr::case_when(
      BestNumber==1 ~ "1",
      BestNumber==2 ~ "2",
      BestNumber==3 ~ "3",
      BestNumber %in% c(4:5) ~ "4:5",
      BestNumber %in% c(6:10)  ~ "6:10",
      BestNumber>10 ~ "> 10"
      ), levels=  c("1","2","3","4:5","6:10","> 10")
  )
)
levels(x$Count) <- c("1","2","3","4:5","6:10","> 10")

table(x$Count) %>% data.frame() %>% csasdown::csas_table(col_names = c("BestNumber","Freq"))

```

### Compare same month of HP sightings in different years

```{r oct-hp-sgt}
coast <- sf::st_read(dsn = "data", layer = "BC_coast_UTM9")
coord <- ggplot2::coord_sf(xlim = c(-125.5, -123), ylim = c(48.1, 49.5), crs = sf::st_crs(4326)) 

# create bathymetric layer
bathy <- getNOAA.bathy(-125.5, -122.5,48.1, 49.5,res=1) %>% 
  fortify(bathy) 
bathy$z[which(bathy$z >= 0)] <- 0
col <- rev(RColorBrewer::brewer.pal(9L, "Blues")[4:7])
col_ramp <- colorRampPalette(col)
#-----------------------------------------------------------

base_map <- ggplot() +
  geom_raster(aes(x=x,y=y,fill = z), data = bathy) +  labs(fill = "Depth (m)") +
  scale_fill_gradientn(colours = col_ramp(20), guide = "none") +
  ggnewscale::new_scale("fill") +
  geom_sf(data = coast, stroke = 0.05, fill = "light yellow", colour = "grey 60") +
  guides(alpha= "none") +
  ylab("")+xlab("") + 
  theme(axis.text.x = element_text(angle = 90))

#-----------------------------------------------------------
# make bathy legend
b_leg <- ggplot() +
  geom_raster(aes(x=x,y=y,fill = z), data = bathy) +  labs(fill = "Depth (m)") +
  scale_fill_gradientn(colours = col_ramp(20)) +
  theme(legend.position = "bottom") +
  guides(fill = guide_colorbar(title.position = "left"))
# leg1 <- gtable_filter(ggplot_gtable(ggplot_build(b_leg)), "guide-box") 
# leg1Grob <- grobTree(leg1)
leg1 <- cowplot::get_legend(b_leg)
#-----------------------------------------------------------
sp <- "Harbour Porpoise"
ap_sf <- all_ap_sf %>% filter(Species %like% sp)

oct_hp <- all_ap_sf %>% filter(month == 10, Species == sp)
oct_lines <- all_effort_lines %>% filter(month == 10)

tx <- oct_hp %>% dplyr::group_by(year) %>%
  dplyr::summarise(N = n()) %>% as.data.frame()

# coord <- ggplot2::coord_sf(xlim = c(-125.4, -123), ylim = c(48.2, 49.44), crs = sf::st_crs(4326)) 

base_map + 
  geom_sf(data = oct_lines, stroke = 0.25, colour = "grey 40")+
  geom_sf(data = oct_hp, stroke = 0.01, alpha = 0.5, aes(size = Count)) +#
  scale_size_manual(values = c(1.5,2,3)) +
  ggtitle(paste0("CeMoRe survey ", sp, " sightings Oct 2020 vs Oct 2021 ")) +
  # annotation_custom(leg1Grob, xmin=-124.8, xmax=-124.95, ymin=47.9, ymax=48.1) +
  coord +
  facet_wrap(~year, drop = F) +
  geom_text(data = tx, aes(x = (-124.4), y = 48.9, label = paste0("N = ", N)), size = 2) 

```

```{r jan-hp-sgt}
jan_hp <- all_ap_sf %>% filter(month == 01, Species == sp)
jan_lines <- all_effort_lines %>% filter(month == 01)

tx <- jan_hp %>% dplyr::group_by(year) %>%
  dplyr::summarise(N = n()) %>% as.data.frame()

base_map + 
  geom_sf(data = jan_lines, stroke = 0.25, colour = "grey 40")+
  geom_sf(data = jan_hp, stroke = 0.01, alpha = 0.5, aes(size = Count)) +#
  scale_size_manual(values = c(1.5,2,3)) +
  ggtitle(paste0("CeMoRe survey ", sp, " sightings Jan 2021 vs Jan 2022 ")) +
  # annotation_custom(leg1Grob, xmin=-124.8, xmax=-124.95, ymin=47.9, ymax=48.1) +
  coord +
  facet_wrap(~year, drop = F) +
  geom_text(data = tx, aes(x = (-124.4), y = 48.9, label = paste0("N = ", N)), size = 2) 


```

### Compare Humpback sightings - winter vs. summer
```{r hw_compare_seasons, echo=FALSE}
sp <- "Humpback"
ap_sf <- all_ap_sf %>% filter(Species %like% sp)
ap_sf$Species %<>% droplevels()
# to order legend symbols consistently
sp <- unique(c(levels(ap_sf$Species)))

hw <- base_map +
  geom_sf(data = all_effort_lines, stroke = 0.25, aes(colour = factor(year)))+
  scale_colour_manual(name = "Effort in year", values = c("pink", "grey 40", "light green")) +
  ggnewscale::new_scale_colour()+
  geom_sf(data = ap_sf, stroke = 0.01, alpha = 0.5, aes(colour = factor(year), size = Count)) +#
  scale_colour_manual("Sightings in year", values = c("red", "black", "green")) +
  scale_size_manual(values = c(1.5,2,3), name = "Number of Individuals") +
  coord

hw +
 ggtitle(paste0("CeMoRe monthly ", sp, " sightings")) +
 facet_wrap(~month_abb, drop = F)
```

### Compare HP sightings - winter vs. summer
```{r hp_compare_seasons, echo=FALSE}
sp <- "Harbour Porpoise"
ap_sf <- all_ap_sf %>% filter(Species %like% sp)
ap_sf$Species %<>% droplevels()
# to order legend symbols consistently
sp <- unique(c(levels(ap_sf$Species)))

hp <- base_map +
  geom_sf(data = all_effort_lines, stroke = 0.25, aes(colour = factor(year)))+
  scale_colour_manual(name = "Effort in year", values = c("pink", "grey 40", "light green")) +
  ggnewscale::new_scale_colour()+
  geom_sf(data = ap_sf, stroke = 0.01, alpha = 0.5, aes(colour = factor(year), size = Count)) +#
  scale_colour_manual("Sightings in year", values = c("red", "black", "green")) +
  scale_size_manual(values = c(1.5,2,3), name = "Number of Individuals") +
  coord

hp +
 ggtitle(paste0("CeMoRe monthly ", sp, " sightings")) +
 facet_wrap(~month_abb, drop = F)
```


Clip by bath layer ~5m to eliminate the shallow areas never surveyable 

Include border 

Plots: 

Remove # transects from summary tables

add se to sightings table

Group size (and other covariates) vs. detection distance
(violin)

Fix extent on maps

Put N on maps

Instead of comparing winter vs. summer, show all months and years (lumped, not by colour), add km sureyed and # years for each. Hours effort?

One fig of full survey plan for one survey with the border, full survey area, diff colours for full vs. Canadian (in Methods? don't worry about coverage scores for now.)

Compare 2 years (PRISMM-Brianna HW JdF,Swiftsure? And/or Anna Hall’s data? And/or Raincoast?) 


Detection functions prelim, Plot  

point form outline of what we plan to show in results
also, scan through PRISMM doc and Thomas' paper for ideas, too