---
title: "dht2"
author: "EK"
date: "11/9/2023"
output: pdf_document
filename: "dht2_hw"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## dht2() comparisions : Humpback data

```{r setup, include=FALSE}

```

# Compare strat_form = surveyid vs year

er_est = R2
innes = F

shows :
winter surveyid, winter year, 
spring surveyid, spring year, 
summer surveyid, summer year,
fall surveyid, fall year

```{r surveyid_v_year include=FALSE}
hw_R2_w_filtered <- readRDS("variance/hw/surveyid/hw_R2_w______filtered.rds")
hw_R2_w_filtered
hw_R2_y_w <- readRDS("variance/hw/year/R2/hw_R2_y_w.rds")
hw_R2_y_w

hw_R2_sp <- readRDS("variance/hw/surveyid/hw_R2_sp.rds")
hw_R2_sp
hw_R2_y_sp <- readRDS("variance/hw/year/R2/hw_R2_y_sp.rds")
hw_R2_y_sp

hw_R2_su <- readRDS("variance/hw/surveyid/hw_R2_su.rds")
hw_R2_su
hw_R2_y_su <- readRDS("variance/hw/year/R2/hw_R2_y_su.rds")
hw_R2_y_su

hw_R2_f <- readRDS("variance/hw/surveyid/hw_R2_f.rds")
hw_R2_f
hw_R2_y_f <- readRDS("variance/hw/year/R2/hw_R2_y_f.rds")
hw_R2_y_f
```

- give SAME Abund, but with year, get lower cv. 
- Makes sense, as with surveyid, transects are grouped by surveyid; with years, 
transects are pooled together within years. So more nested rep's within each year than 
were within each surveyid (all transects from all surveys within a season year pooled).

# Compare r2 v o2

strat_form = year,
innes = F

shows :
winter r2, winter o2, 
spring r2, spring o2, 
summer r2, summer o2,
fall r2,     fall o2

```{r r2_v_o2}
hw_R2_y_w
hw_O2_y_w <- readRDS("variance/hw/year/O2/hw_O2_y_w.rds")
hw_O2_y_w

hw_R2_y_sp
hw_O2_y_sp <- readRDS("variance/hw/year/O2/hw_O2_y_sp.rds")
hw_O2_y_sp

hw_R2_y_su
hw_O2_y_su <- readRDS("variance/hw/year/O2/hw_O2_y_su.rds")
hw_O2_y_su

hw_R2_y_f
hw_O2_y_f <- readRDS("variance/hw/year/O2/hw_O2_y_f.rds")
hw_O2_y_f

```

```{r cars}
summary(cars)
```



```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
